Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-03 summary.tex
rough version 2 done

\section{Summary and discussion}\label{summary}
%what have we done, relation to previous work 
In this paper we have studied the linear axisymmetric vertical shear
instability (VSI) in astrophysical disks that arises from the
height-dependence of the equilibrium angular velocity profile
$\Omega(r,z)$ of a baroclinic disk. The VSI was analyzed and
explained earlier in the local limit by \cite{urpin98} and
\cite{urpin03}; and simulated by \cite{arlt04} and more recently by
\cite{nelson13}.

We generalized previous vertically-global linear stability calculations
of the VSI \citep{nelson13,mcnally14}, which assumed isothermal
perturbations, to include an energy equation. Our effort may also be
regarded as an extension of \cite{urpin03}, which included an energy
equation, by relaxing the Boussinesq and vertically-local
approximations adopted in that analysis.

We added a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale $t_c=\beta\Omega_k^{-1}$. This
form of thermal relaxation was also employed in the global numerical
simulations of \cite{nelson13}, and allows the thermodynamic response
of the disk to be controlled.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially-local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}, who showed the linear VSI studied by
\cite{nelson13} in a locally isothermal disk can be captured by the
VGSB. We find the VGSB framework is appropriate for $\beta\lesssim1$. 
This is convenient because previous works show that rapid thermal
relaxation ($\beta\ll1$) is generally required for the VSI. However,
the VGSB framework is not valid in the adiabatic limit.

@@

We mostly focused on vertically isothermal disks ($\Gamma=1$) because
it permits a largely analytical discussion. Our analysis is an
extension of the linear models considered in \cite{lubow93} by
adding vertical shear and (rapid) thermal relaxation, but in the limit of low-frequency
perturbations in nearly-Keplerian disks. These approximations were
made for simplicity, but also motivated by previous work
\citep{nelson13}.

For isothermal perturbations ($\beta=0$) we explicitly
demonstrated instability  associated with vertical shear, and that the 
VSI growth rate is limited by the maximum vertical shear within the 
vertical domain of interest. The fundamental VSI mode is characterized
by one node in the density perturbation at the mid-plane
(i.e. anti-symmetric about $z=0$), which corresponds to a vertical
velocity perturbation  $\delta v_z\sim\mathrm{constant}$ near
$z=0$.

% In the opposite limit of adiabatic perturbations ($\gamma>1$
% and $\beta\to\infty$), we find VSI growth rates are dramatically
% reduced on account of a stabilizing 
% vertical entropy gradient. The VSI is not completely
% surpressed because vertical shear dominates over buoyancy 
% sufficiently close the disk midplane. However, perturbations grow
% slowly and is constrained to regions immediately about the disk
% midplane. Thus, the adiabatic VSI is likely unimportant in practice.

For non-isothermal  perturbations with thermal relaxation ($\gamma>1$,
$\beta>0$) , the general dispersion relation (Eq. \ref{relax_disp}) is
complicated and requires a numerical treatment. However, for the 
fundamental VSI some simplification is  possible. In particular, we
quantified the critical thermal relaxation timescale
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below which small-scale
perturbations may be regarded as isothermal and the fundamental VSI
operates.  We expect $\beta<\beta_\mathrm{crit}$ to be a sufficient
condition for the VSI to operate. We believe this explicit expression for
$\beta_\mathrm{crit}$ is the most useful result from our
investigation.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we accounted for
the background radial disk structure for consistency. We focused
on the fundamental mode to compare with our analytical
discussion, and because \cite{nelson13} found that this mode 
eventually dominate their simulations. Our numerical solutions are
mostly consistent with our analytic expectations, including the 
dependence of $\beta_\mathrm{crit}$ on disk parameters.

@@

We applied our linear results to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae (MMSN),
using disk models described in \cite{chiang10}. We find the MMSN is
unstable to the fundamental VSI at $10$s of AU, with characteristic
growth times of $10$s of orbits. However, we note the strong
temperature dependence of the physical thermal relaxation timescale
in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
radial scale). Thus, increasing the disk temperature even slightly can significantly 
shorten the typical growth times for the VSI. This is also attributed
to a larger value of $\beta_\mathrm{crit}\propto\epsilon$, since increasing $T$ also
increases the aspect-ratio $\epsilon$.

@@

\subsection{Caveats and outlooks} 
We highlight below some improvements to the present  
linear models in future works:

\emph{Realistic thermal evolution.} We have treated the thermal
relaxation time $t_c$ as a constant for simplicity. However, a
physical expression for $t_c$ will depend on 
$z$. For example, the Newtonian cooling timescale
(\S\ref{newton_cool}) will depend on $z$ for vertically non-isothermal
disks, and through the opacity since 
$\kappa_d=\kappa_d(\rho,T)$ in general. 
A proper treatment for radiative diffusion, on the other hand, will
ultimately involve the Laplacian operator. In this case additional
boundary conditions must be supplemented. We will consider this in a
future work.

%need to compare kley sims

@@

\emph{Radial dependence.} We have adopted the radially-local approximation 
throughout this paper. This is based on previous works, including
nonlinear simulations, which show that the VSI is relevant for
perturbations with small radial lengthscales. Although this simplifies
the linear problem to ordinary differential equations, this approach
introduces the wavenumber $k_x$ as a free parameter. The effect
of radial boundary conditions cannot be explored. Future
generalizations of the linear problem should retain the radial
dependence properly and solve the two-dimensional $(r,z)$ partial 
differential equation.

>>>>

\section{Summary and discussion}\label{summary}
%what have we done, relation to previous work 
In this paper we have studied the linear axisymmetric vertical shear
instability (VSI) in astrophysical disks that arises from the
height-dependence of the equilibrium angular velocity profile
$\Omega=\Omega(r,z)$ in a baroclinic disk. We examined how
the linear instability depends on disk thermodynamics and applied our
results to assess the relevance of the VSI in realistic protoplanetary
disks.

% The VSI was analyzed earlier in 
% the local limit by \cite{urpin98} and 
% \cite{urpin03} and simulated by \cite{arlt04}. Recent non-linear
% numerical simulations by \cite{nelson13} and \cite{stoll14} indicate
% the VSI generally requires a short thermal timescale.

We generalized previous vertically global linear stability calculations
of the VSI \citep{nelson13,mcnally14}, which assumed isothermal
perturbations, to include an energy equation. Our effort may also be
regarded as an extension of \cite{urpin03}, which did include an
energy equation, by relaxing the Boussinesq and vertically local
approximations adopted in that analysis.

We included a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach to model thermal relaxation was also
employed in the global numerical simulations of \cite{nelson13}, and
allows the thermodynamic response of the disk to be controlled.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}, who showed the linear VSI studied by
\cite{nelson13} in a locally isothermal disk can be captured by the
VGSB. We find the VGSB framework is appropriate for $\beta\lesssim1$. 
This is convenient because rapid thermal relaxation ($\beta\ll1$) is
generally required for the VSI.  However,
the VGSB framework is not appropriate in the adiabatic limit ($\beta \gg
1$).

@@

We mostly focused on vertically isothermal disks ($\Gamma=1$), since
this is the relevant case for protoplanetary disks that are irradiated
\citep{chiang97}. It also permits a largely analytical discussion.  Our analysis is an
extension of the linear models considered in \cite{lubow93} by 
adding vertical shear and (rapid) thermal relaxation, but in the limit of low-frequency
perturbations in nearly-Keplerian disks. These approximations were
made for simplicity, but also motivated by recent numerical
simulations \citep{nelson13}.

For isothermal perturbations ($\beta=0$) we explicitly
demonstrated instability  associated with vertical shear, and that the 
VSI growth rate is limited by the maximum vertical shear within the 
vertical domain of interest. The fundamental VSI mode is characterized
by one node in the density perturbation at the midplane 
(i.e. anti-symmetric about $z=0$), which corresponds to a vertical
velocity perturbation  $\delta v_z\sim\mathrm{constant}$ near the
mid-plane with zero nodes. These have been referred to as
`corrugation modes' by \cite{nelson13}, and were found to dominate
their numerical simulations, as well as that of
\cite{stoll14}. Furthermore, when computing unstable modes for a
disk with a physical model for thermal relaxation, we also find
the fundamental mode to be dominant and robust. Such modes are thus
likely most relevant to  realistic disks.

For non-isothermal  perturbations with finite thermal relaxation ($\gamma>1$,
$\beta>0$) , the general dispersion relation (Eq. \ref{relax_disp}) is
quite complicated. However, for the 
fundamental VSI some simplification is  possible. In particular, we
quantified the critical thermal relaxation timescale
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below which small-scale
perturbations may be regarded as isothermal and the fundamental VSI
operates.  We expect $\beta<\beta_\mathrm{crit}$ to be a sufficient
condition for the VSI to operate. We have therefore provided a
quantitative measure for the`fast' thermal relaxation requirement for the
VSI.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we relax all the
approximations made in our analytical discussion, and accounted for
the background radial disk structure for consistency.  Our
numerical solutions are mostly consistent with our analytic 
expectations, including the dependence of $\beta_\mathrm{crit}$ on
disk parameters.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae
(MMSN). We consider the disk model described in \cite{chiang10}, and
modeled the thermal relaxation timescale $\beta$ based on
dust opacity, which results in a $\beta$ that depends on both the
lengthscale of interest as well as the height from the midplane.  
We find the MMSN is unstable to the fundamental VSI at $10$s of AU,
with characteristic growth times of $\sim 30$ orbits. Although $\beta$
is not spatially constant as assumed in our analysis, we find that
evaluating $\beta < \beta_\mathrm{crit}$ at the disk
midplane gives a rough measure of the radii at which the VSI operates in the
MMSN.  This simple criteron may be a useful first step to 
estimate the relevance of the VSI in realistic astrophysical
disks without resorting to non-linear numerical simulations.

%  However, we note the strong
% temperature dependence of the physical thermal relaxation timescale
% in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
% radial scale). Thus, increasing the disk temperature even slightly can significantly 
% shorten the typical growth times for the VSI. This is also attributed
% to a larger value of $\beta_\mathrm{crit}\propto\epsilon$, since increasing $T$ also
% increases the aspect-ratio $\epsilon$.

@@

\subsection{Caveats and outlooks} 
We highlight below some extensions to the present  
linear models in future works:

\emph{Improved thermal evolution.} We assumed non-adiabatic effects
can be represented by a thermal relaxation timescale $t_c$. For 
application to protoplanetary disks we connected $t_c$ to the disk
properties, but one can improve the realism by generalizing the
opacity model adopted here to be density-dependent,  i.e. $\kappa_d =
\kappa_d(\rho, T)$.  However, going beyond the framework of thermal
relaxation requires a proper treatment for radiative diffusion, which
involves the Lapacian operator and additional boundary conditions
must be supplemented. We will consider these improvements in a
follow-up study.

@@

\emph{Radial dependence.} Based on previous works, including
non-linear simulations, which show that the VSI is relevant for
perturbations with small radial lengthscales, we have adopted the
radially-local approximation throughout this paper.  Although this
simplifies the linear problem to ordinary differential equations, this
approach introduces the wavenumber $k_x$ as a free parameter. The
effect of radial boundary conditions cannot be explored. Future 
generalizations of the linear problem should retain the radial
dependence properly and solve the two-dimensional $(r,z)$ partial 
differential equation eigenvalue problem.
